---
title: "part2-analysis.qmd"
author: "coletterouiller"
format: html
jupyter: /opt/homebrew/Caskroom/miniconda/base/bin/python
editor: visual
---

# Part 2: Fatty Liver Disease Classification

## Background

Fatty Liver Disease (FLD) is a liver condition characterized by the accumulation of lipid droplets in the liver, affecting up to 42.2% of adults in the US (Jones et. al, 2022). FLD causes impaired liver function; though it is often asymptomatic, it can include symptoms like abdominal pain/discomfort, weight loss, loss of appetite, and fatigue (Zhu, 2024). The Hepatic Steatosis Index is a screening tool used to diagnose Non-Alcoholic Fatty Liver Disease (NAFLD), which is a subsection of FLD associated with excessive alcohol consumption (Jones et. al, 2022). Since HSI was not included in the initial data sets but is calculated formulaically, I will be using it as the ground truth label for each patient.

HSI is calculated by:

*𝐻𝑆𝐼* = 8 ∗ (*𝐴𝐿𝑇* /*𝐴𝑆𝑇* ) + *𝐵𝑀𝐼* + (2, *𝑐𝑜𝑛𝑑* : *𝑑𝑖𝑎𝑏𝑒𝑡𝑒𝑠*) + (2, *𝑐𝑜𝑛𝑑* : *𝑓𝑒𝑚𝑎𝑙𝑒*)

The following is a description of each variable/it's importance in the HSI formula:

1\) Alanine Aminotransferase (ALT) = liver enzyme used as a biomarker for liver cell injury/damage (Lee et. al, 2009).

2\) Aspartate Aminotransferase (AST) = tissue enzyme that is used as a biomarker for tissue injury/damage since it is found in muscle, skeletal, neural, cardiac,and hepatic tissue (Lee et. al, 2009).

When hepatocytes are functioning abnormally/exhibiting stress, ALT rises more than AST, so elevated ALT/AST ratio is known to suggest liver damage/oxidative stress consistent with FLD (Lee et. al, 2009).

3\) BMI = body composition metric calculated by body weight (kg)/height\^2 (m\^2) (Lee et. al, 2009).

4\) Sex = Sex is relavent due to the fact that adipose tissue distribution, hormone levels, and enzyme activity di"ers between male and female (Lee et. al, 2009).

5\) Diabetes = Type II diabetes is strongly linked to FLD. Therefor it is included for diagnostic relevency (Lee et. al, 2009).

'''SECTION NEEDS TO RE-TAILORED" Then I included multiple additional features that are associated with FLD: homocysteine, serum folate, red blood cell (RBC) folate, (MMA), (GGT), C-Reactive Protein (CRP), triglyceride, Ferritin, and Vitamin B12 levels.

1\) Homocysteine = a folate metabolism byproduct that is created when methioinine, an amino acid necessary for all methylation pathways, donates a methyl group (Martinez et al, 2017). Homocysteine is not inherently harmful, but elevated levels have been associated with FLD (Ma!é et. al, 2009). Case-controlled studies have shown that NAFLD patients show significantly higher homocysteine levels than controls (DeCarvalho, 2013).

2)  Serum Folate = folate de\$ciency impares methylation and hepatic lipid metabolism. Therefore, we expect and studies have shown that NAFLD patients have displayed low serum folate levels compared to control groups (Liu et. al, 2023).

3)  Vitamin B12 = is an important cofactor for methylation and carbon metabolism (Jones et. al, 2022). The CDC has noted Vitamin B12 levels differ between NAFLD and controls, as per the National Health and Nutrition Examination Survey (NHANES) 1999-2004 cohort analysis (Li et. al, 2022).

...

The model will use these additional features to determine class assignment, and then compare to the HSI ground truth we established to determine model accuracy. The data used for this model comes from 2001-2002 and 2003-2004 cohorts of the National Health and Nutrition Examination Survey (NHANES) from the CDC.

### Define Task

Task: Build an ML model that uses the features extracted from the NHANES data set labeled to predict whether a patient has Non-Alcoholic Fatty Liver Disease (NAFLD).

This task is an example of supervised learning because it is a classification task, which is a sub-type of supervised machine learning. Since I only have 2 classifications for the model (NAFLD or Healthy) this is a binary classification.

## Prepare Data

### Load Data 

The code block below loads my dataset, which contains one row per patient. For each patient, I have multiple measured features including : ALT, ALS, BMI, sex, homocysteine, serum folate, red blood cell (RBC) folate, (MMA), (GGT), C-Reactive Protein, (CRP), triglyceride, Ferritin and B12 levels, as well as BMI, sex and diabetic status.

Each cell also has a label for HSI score that represents the true class of the cell, with HSI scores greater or equal to 36 being the determinate threshold score.

```{r}
##python was acting unstable and automatically using the reticulate cache - created .Renviron file to fix this issue so the same python and interpretere are used everytime RStudio or Quarto runs. 

##for reproducibility purposes, the export package list can be found in the requirements.txt file. 

##first check
Sys.getenv("RETICULATE_PYTHON")

##second check 
library(reticulate)
py_config()
```

```{python}
import sys
print(sys.executable)
```

```{python}

# Load libraries and dataset

import graphviz 
import pandas as pd
import numpy as np
import seaborn as sns
from sklearn.model_selection import KFold
from sklearn.model_selection import train_test_split
from sklearn.neighbors import KNeighborsClassifier
from sklearn.linear_model import LogisticRegression
from sklearn.tree import DecisionTreeClassifier, plot_tree
from sklearn.metrics import accuracy_score, confusion_matrix
from sklearn.preprocessing import StandardScaler
import matplotlib.pyplot as plt
from sklearn.model_selection import cross_val_score
import graphviz
from sklearn.tree import export_graphviz
from io import StringIO
from IPython.display import display

#set to 42 to ensure training/test spits are reproducible 
MY_RANDOM_STATE = 42

# Load NHANES cleaned data set 
df = pd.read_csv("../../../data/clean_data/nhanes_final.csv")

```

```{python}

#check data
print(df.shape)
print(df.columns)
print(df.dtypes)

#check missingness 
print(df.isna().sum())

#binary check 
df[['alt','ast','bmi','homocysteine']].describe()

```

These results look good to proceed. Datatypes are correct, core variables are loaded correctly, and there are no obvious signs of data corruption. It is important to note that these checks revealed a moderate-high level of missingness in the biomarker variables. This information was first discovered in preprocessing and therfore cohorts were not defined by full overlap. There appears to be some extreme outliers that will need to be addressed downstream. There may also need to be some encoding adjustments across variables sex (int64), race_ethnicity (int64) and cohort (object) downstream as well.

### Adjust Variables 

```{python}
#need to set/extract relevant label values
#ALT/AST ratio
df['alt_ast_ratio'] = df['alt'] / df['ast']

#female flag: 1 = female, 0 = male
df['female'] = (df['sex'] == 2).astype(int)

#diabetic flag: # 1 = yes, 0 = no
df['diabetic'] = (df['diabetes'] == 1).astype(int)

#compute HSI and store as new variable
#formula: HSI= 8*(ALT/AST)+BMI+(2, if diabetes)+(2, if female)
df['HSI'] = (
    8 * df['alt_ast_ratio']
    + df['bmi']
    + 2 * df['diabetic']
    + 2 * df['female']
)

#new use HSI to determine NAFLD label (binary ML label)
#scoring: 1 = NAFLD (HSI >= 36), 0 = healthy
df['NAFLD'] = (df['HSI'] >= 36).astype(int)

#three-level interpretive category (used downstream)
df['HSI_category'] = np.where(
    df['HSI'] < 30, "healthy",
    np.where(df['HSI'] < 36, "borderline", "nafld_like"))


```

In this section, I corrected the encoding for the sex and diabetes values used in the HSI calculations and established a structure that creates three new variables: HSI, NAFLD, and HSI_category. HSI functions as a continuous score, and it's derived variable, NAFLD, acts as the primary binary ML label. HSI_category was created to provide a descriptive severity grouping for downstream analysis in R, which includes the groups: healthy, borderline, and nafld_like.

### Interpret and Check Data 

The following code below displays the first few rows of the data set and views summary statistics for the HSI, NAFLD, and HSI_category labels.

```{python}
# 1. Display first few rows
print("\n1. Sample records:\n")
display(df.head())
print("\n")

## 2. Summary statistics for features
print("\n2. Summary statistics:\n")
#HSI distribution 
print(df['HSI'].describe())

#HSI category counts 
print(df['HSI_category'].value_counts())

#ML label balance 
print(df['NAFLD'].value_counts())

#cross check bins 
print(pd.crosstab(df['HSI_category'], df['NAFLD']))
```

Through these checks, we can see the HSI distribution mean is 35.47 and the median is 34.53, which is aligned with our expectations given the 36.0 value cut off, confirming an absense of scaling issues or computation errors. The categorical distribution for downstream analysis is also promising, with 'nafld-like', 'borderline', and 'healthy' showing 6,977, 4022, and 3,416 counts, respectively. These counts are promising because it showed all three bins present without being dominated by one class.

The cross tab check evaluates our binning method for logical inconsistencies. The 'borderline' and 'healthy' bins behave as expected, with 0 individuals being labeled as 'NAFLD' = 0. However, the 'nafld_like' bin results show that not all individuals in the high HSI range are classified as NAFLD-positive under the strict established threshold.

Functionally, this means that while all NAFLD-positive fall into the 'nafld_like' bin, a subset of individuals remain below the diagnostic threshold that still qualify as "high risk." This inconsistancy accurately reflects HSI as a continuous metric and proxy measure for FLD diagnosis, which supports my analysis approach to include a 'borderline' bin.

### Feature Definition 

My input features are have 14 measured features: ALT, ALS, homocysteine, serum folate, red blood cell (RBC) folate, (MMA), (GGT), C-Reactive Protein, (CRP), triglyceride, Ferritin and B12 levels, as well as BMI, sex and diabetic status. My labels are '1' for NAFLD and 0 for 'Healthy', which correspond to the 'NAFLD' column of our dataset.

Because the variables containing biomarker data is limited in comparison to the HSI variables, I will need to create mutliple models. Each model will have their own features, datasets, and set x and y values.

```{python}

#define features (x) and label (y)

# model-specific features 
#core/baseline features (HSI)
core_feature_columns = [
    'alt',   # ALT
    'ast',   # AST
    'bmi',     # BMI
    'female',     # 0/1
    'diabetic'    # 0/1
]

#extended features (biomarkers)
bio_feature_columns = [
    'homocysteine',   # ALT
    'serum_folate',   # serum folate
    'rbc_folate',     # rbc folate
    'crp',     # crp
    'mm_acid',    # mma
    'ggt', #ggt
    'vit_b12', #vitamin b12
    'ferritin', #iron
    'alt',   # ALT
    'ast',   # AST
    'bmi',     # BMI
    'female',     # 0/1
    'diabetic'    # 0/1
]

#model-specific data sets 
df_core = df[core_feature_columns + ['NAFLD']].dropna()
df_bio = df[bio_feature_columns + ['NAFLD']].dropna()

#now set feature/label
#core
X_core = df_core[core_feature_columns]
y_core = df_core['NAFLD'] # 0 = healthy, 1 = NAFLD

#biomarker
X_bio = df_bio[bio_feature_columns]
y_bio = df_bio['NAFLD'] # 0 = healthy, 1 = NAFLD

#now check sample sizes 
print("Baseline model shape:", df_core.shape)
print("Biomarker model shape:", df_bio.shape)


#now check class balances 
print(y_core.value_counts())
print(y_bio.value_counts())


#quick sanity check of inputs
X_core.head(), X_bio.head()
```

This code chunk shows that we retained nearly 13,000 observations for the core feature set and about 2,500 for the biomarker feature set. While not ideal, this feature shows that richer biological detail often comes at the cost of sample size.

The class balances are the same at \~57/43 split, which is a healthy split for classification. It is important to note that this consistency suggests that missingness did not distort class balance, and therefor the complete-case biomarker subset is not obviously biased towards a single outcome class. The biomarker model therefore represents a smaller but biologically richer subset rather than a distorted sample.

-   stopping point

```{r}
1 + 1


```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
