---
title: "2.3 Visualization & Analysis"
format: 
  html:
    css: styles.css
    code-fold: false
    code-summary: "Show code"
toc-depth: 5
---

### <u> 2.3.1: Overview & Setup </u>

The R portion of Part 2 functions as a diagnostic and interpretive layer on top of the machine learning pipeline established in section 2.2, which focused on classification modeling and validation metrics.

##### <u> 2.3.1.1: Purpose </u>

This section uses R to examine the structural and distributional features of the dataset that shaped the ML portion results. R was selected for this portion because it is often used by computational biologists for cohort stratification and distributional trend analysis, which aligns with the core analytic approach for this stage of the project.

Specifically, this section addresses four questions left open or only partially resolved by the ML Python analysis:

1.  How is missingness structured in the biomarker subset, and how well does that subset demographically represent the full cohort?

2.  Do the novel biomarker distributions separate meaningfully across HSI severity categories, and do those distributions contribute to an explanation for the counter-intuitive logistic regression coefficients observed in section 2.2.4.3?

3.  What are the structural implications of the sex imbalance identified in the biomarker subset section 2.2.4.4 for downstream interpretation?

These diagnostic questions are designed to complement the modeling work in section 2.2, rather than replicate it, so there is no models refitting present in the R analysis. Rather, the intent of this portion is to investigate the dataset characteristics that contextualize Part 2's findings, especially those that remain unclear, to improve the weight and clarity of the coding contribution in Part 1–Part 2 integration discussion.

##### <u> 2.3.1.2: Load Packages & Data </u>

Packages are loaded using "pacman::p_load" to streamline installation.

The dataset that is read in below is the final analysis-ready NHANES file that was constructed via the PostgreSQL pipeline described in section 2.1 and used for the Python analysis in section 2.2.

Full scripts for file construction pipeline can be found on the GitHub project repository (link: https://github.com/colette-osiris/mthfr-fld-capstone).


```{r}
#| label: install-packages 
#| code-fold: true
##| output: false


# LOAD PACKAGES -  -------------
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse,
               stringr,
               ggridges, #potential use for future data analysis 
               janitor, #new feature 
               viridis, 
               patchwork, 
               scales, 
               here, 
               grid, 
               flextable, 
               update = FALSE)
```

```{r}
#| label: data-import-exploration 
#| code-fold: true
##| output: false

# read in final analysis-ready NHANES file from PostgreSQL pipeline (sect. 2.1)
NHANES.dataset.01 <- readr::read_csv(here::here("data",
                                        "clean_data",
                                        "nhanes_final.csv"),
                                        name_repair = "universal")

##explore data 
#glimpse(NHANES.dataset.01)
#names(NHANES.dataset.01)
cat("Loaded: ", nrow(NHANES.dataset.01), "rows and", ncol(NHANES.dataset.01), "columns.")

```

##### <u> 2.3.1.3: Label Reconstruction </u>

HSI and its derived labels (NAFLD, HSI_category) are re-constructed here rather than imported from the Python analysis for reproducibility purposes. This way, any reader running this file in isolation arrives at the same labels that were used in section 2.2 independently.

```{r}
#| label: HSI-component-derivatives 
#| echo: true
#| output: false


# DERIVE HSI COMPONENTS & LABELS -  -------------
#
#using same formulas as 2.2.3.4 in ML doc
#adapt from project code - look at PLSC426 content for exact formatting 
NHANES.dataset.01 <- NHANES.dataset.01 %>%
  dplyr::mutate(
    #sex indicator: 1 = female, 0 = male
    female = dplyr::case_when(
      sex == 2 ~ 1L, 
      TRUE ~ 0L),
    #diabetes indicator: 1 = yes, 0 = no
    #need to collapse all other options (7/9/etc)
    diabetic = dplyr::case_when(
      diabetes == 1 ~ 1L, 
      TRUE ~ 0L),
    #ALT/AST ratio (used in HSI)
    alt_ast_ratio = alt / ast,
    #HSI formula: 8*(ALT/AST) + BMI + 2*(if diabetic) + 2*(if female)
    HSI = 8 * alt_ast_ratio + bmi + (2 * diabetic) + (2 * female),
    #binary NAFLD label: 1 = NAFLD (HSI >= 36), 0 = healthy
    NAFLD = as.integer(HSI >= 36),
    #same three-level severity category - use later for distributional analysis
    HSI_category = dplyr::case_when(
      #need to address NA so they dont get misclassified 
      is.na(HSI) ~ NA_character_,
      HSI < 30 ~ "healthy",
      HSI < 36 ~ "borderline",
      HSI >= 36 ~ "nafld_like"), #don't forget TRUE catch-all
    #set ordered factor so healthy -> borderline -> nafld_like reads left-to-right
    HSI_category = forcats::fct_relevel(HSI_category,
                                        "healthy", 
                                        "borderline", 
                                        "nafld_like"))

#view it 
NHANES.dataset.01
```

Because of the reproducibility practices used, the following chunk checks the data distribution in order to ensure R-derived labels stay consistent throughout the analysis.  

```{r}
#| label: HSI-category-flextable 
#| code-fold: true
#| echo: true
#| output: true

#confirm category counts match Python-derived labels from §2
hsi_category_flextable <- NHANES.dataset.01 %>%
  dplyr::count(HSI_category, 
               name = "count") %>%
  #use flextable formatting from class 
  flextable::flextable() %>%
  flextable::theme_booktabs(bold_header = TRUE) %>%
  flextable::align(align = "center", part = "all") %>%
  flextable::valign(valign = "bottom", part = "header") %>% 
  flextable::padding(padding = 1, part = "all") %>% 
  flextable::autofit()


#view it 
hsi_category_flextable
```


```{r}
#| label: HSI-category-flextable-save
#| code-fold: true
#| echo: false
#| output: false

#save flextable 
flextable::save_as_image(x = hsi_category_flextable, path = "../plots/part2/hsi_category_flextable.png")
```

The category counts (healthy = 3,416, borderline = 4,022, nafld_like = 5,551) match those found in the ML analysis, which confirms the R derivation reproduces the Python labels correctly. This integrity check was initially included in the portion of the project to act as a routine reproducibility verification measure. However, during preliminary runs, it revealed a discrepancy in the ML code that silently misclassified 1,446 rows with missing HSI inputs. This error was traced to the Python code's default NaN behavior in section 2.2.3.4 and was resolved, along with all downstream implications of the initial bug. With labels reconstructed and validated, the analysis proceeds to the diagnostic examinations outlined below.

### <u> 2.3.2: Missingness & Subset Representativeness </u>

##### <u> 2.3.2.1: Biomarker Missingness by Cohort </u>

Biomarker subset structures are first explored by examining missingness across both NHANES cohorts. This was chosen as the first step since missingness directly shapes what observations th model can use and therefore could distort both the subset used in downstream analysis and any accompanying interpretation.

```{r}
#| label: full-missingness-table
#| code-fold: true
#| output: true

# BIOMARKER MISSINGNESS HEATMAP -  -------------
#tile heatmap of biomarker coverage % per cohort 
full_missingness_data <- NHANES.dataset.01 %>% 
  dplyr::select(cohort,
                alt,
                ast, 
                diabetes, 
                bmi,
                sex, 
                homocysteine, 
                serum_folate, 
                rbc_folate, 
                crp,
                mm_acid, 
                ggt, 
                vit_b12, 
                ferritin) %>% 
  #pivot longer so biomarkers become a variable 
  tidyr::pivot_longer(
    cols = -cohort, 
    names_to = "biomarker",
    values_to = "value") %>% 
  #compute coverage per biomarker per cohort 
  dplyr::summarise(
    coverage_pct = sum(!is.na(value)) / dplyr::n() * 100,
    .by = c(cohort, biomarker)) %>% 
  dplyr::mutate(
    #rename biomarkers for cleaner display 
    biomarker = dplyr::recode(
      biomarker,
      "homocysteine" = "Homocysteine",
      "serum_folate" = "Serum Folate",
      "rbc_folate" = "RBC Folate",
      "crp" = "CRP",
      "mm_acid" = "MMA",
      "ggt" = "GGT",
      "vit_b12" = "Vitamin B12",
      "ferritin" = "Ferritin",
      "alt" = "ALT",
      "ast" = "AST",
      "bmi" = "BMI",
      "sex" = "Sex",
      "diabetes" = "Diabetes"),
    #order biomarkers by lowest avg coverage 
    biomarker = forcats::fct_reorder(biomarker, coverage_pct, .fun = mean),
    #clean cohort labels 
    cohort = dplyr::case_when(
      cohort == "c1_2001_2002" ~ "2001-2002",
      cohort == "c2_2003_2004" ~ "2003-2004"))

#plot 
full_missingness_plot <- full_missingness_data %>% 
  ggplot(aes(
    x = cohort, 
    y = biomarker, 
    fill = coverage_pct)) +
  geom_tile(color = "white") + 
  #add % labels in cells - had to look up correct geom for annotations like this
  geom_text(
    aes(label = paste0(round(coverage_pct, 1), "%")),
    color = "white", #changed - black looked odd 
    fontface = "bold",
    size = 4) +
  #color theme is same as R project but continous 
  scale_fill_viridis_c(
    option = "plasma",
    begin = 0, 
    end = 0.7,    #avoid blinding yellow end 
    name = "Coverage (%)") +
  labs(
    x = "Cohort", 
    #changed label from biomarker to feature
    y = "Feature",
    title = "Feature Coverage by NHANES Cohort \n (% of Participants with Non-Missing Values)") +
  theme_minimal() +
  #title/legend position wrangling 
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank())

#save plot - ggplot not flextable syntax 
#ggsave(
  #filename = here::here("plots", "part2", "biomarker_missingness_tile.png"),
  #plot = missingness_plot,
  #dpi = 400,
  #width = 10,
  #height = 7,
  #units = "in",
  #bg = "white") #was saving grey - this fixed it

#view it 
full_missingness_plot
```

```{r}
#| label: full-missingness-table-save
#| echo: false
#| output: false

#save plot - ggplot not flextable syntax 
ggsave(
  filename = here::here("plots", "part2", "biomarker_missingness_tile.png"),
  plot = full_missingness_plot,
  dpi = 400,
  width = 10,
  height = 7,
  units = "in",
  bg = "white") #was saving grey - this fixed it
```

The most notable feature of the figure is the sharp coverage collapse of six biomarkers in cohort 1. This is a significant cohort imbalance, though it's notable that uniform distribution of affected variables across this skew stays relatively stable (7.3-7.4%). This pattern more likely reflects a source-level NHANES data artifact rather than a pipeline error, given that the ~7% coverage matches the documented biomarker subset size from preprocessing in section 2.1.5. 

Additionally, it is worth noting that the only two biomarkers unaffected by the coverage collapse are CRP and GGT, which each show coverage of ~93% across both cohorts. The data for those biomarkers did not come from the same source file as the affected biomarkers (L06_2_B.XPT), which accounts for the coverage discrepancy. Similarly, ferritin in cohort 2 appears to have a 31.3% coverage, which matches the preprocessing value in section 2.1.7. Like CRP and GGT in cohort 1, ferritin data for cohort 2 was stored in a separate document, further indicating that feature coverage is directly impacted by source filing mechanisms used by NHANES. 

Overall, this missingness plot shows that cohort 1 has a severe biomarker gap from the L06 source file structure, while cohort 2 has more uniform coverage, with the exception of ferritin. This coverage structure directly shapes the biomarker subset that was used in downstream ML analysis since the model can only train on observations with non-missing values across all biomarkers. This restricts the analytical sample to 7.4% and 31.3% of initial participants of cohort 1 and 2, respectively, which explains why the biomarker model was trained on ~2,500 observations while the core model was trained on ~13,000 observations. This answers the first part of the question: 'how is missingness structured in the biomarker subset, and how well does that subset demographically represent the full cohort?', though further analysis is necessary to determine how well the subset represents the full cohort. 

##### <u> 2.3.2.2: Demographic Comparison: Core vs Biomarker Subset </u>

This section explores demographic comparisons between the two subsets, which is important to consider for generalizability of any interpretive conclusions. Section 2.2.4.4 established a strong sex skew within the biomarker model (89% female vs 51% in the full cohort) that functions as a key limitation and requires further investigation. 

```{r}
#| label: core-vs-biomarker-dataset-extraction
#| code-fold: true
#| echo: true
#| output: true


# DEMOGRAPHIC COMPARISON: CORE VS BIOMARKER SUBSET -  -------------
#want to compare sex and race/ethnicity across full core dataset and biomarker subset
#to show how selection bias manifests in the complete-case subset 

#first need to set up both datasets 
#core dataset - same columns Python uses for core model (drop NAs in core vars)
core_dataset <- NHANES.dataset.01 %>% 
  dplyr::filter(
    !is.na(alt), 
    !is.na(ast), 
    !is.na(bmi), 
    !is.na(sex), 
    !is.na(diabetes)) %>% 
  dplyr::mutate(dataset = "Core Dataset")

#biomarker subset - all 13 features (8 biomarkers + 5 core)
biomarker_subset <- NHANES.dataset.01 %>% 
  dplyr::filter(
    !is.na(homocysteine), 
    !is.na(serum_folate), 
    !is.na(rbc_folate), 
    !is.na(crp), 
    !is.na(mm_acid), 
    !is.na(ggt), 
    !is.na(vit_b12), 
    !is.na(ferritin), 
    !is.na(alt), 
    !is.na(ast), 
    !is.na(bmi), 
    !is.na(sex), 
    !is.na(diabetes)) %>% 
  dplyr::mutate(dataset = "Biomarker Subset")

#print sample sizes for prose reference 
cat("Core dataset n:", nrow(core_dataset), "\n")
cat("Biomarker subset n:", nrow(biomarker_subset), "\n")
```


```{r}
#| label: core-vs-biomarker-demographic-plot-prep
#| code-fold: true
#| output: true

#stack them into one long df for plotting - look at data for actual labels used
demographic_comparison <- dplyr::bind_rows(core_dataset, biomarker_subset) %>% 
  dplyr::mutate(
    #recoded sex for readability 
    sex_label = dplyr::case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female"),
    #recode race/ethnicity using NHANES codes - need to look at data for actual used
    #labels found as follows: 
    #1 = Mexican American, 2 = Other Hispanic, 3 = Non-Hispanic White,
    #4 = Non-Hispanic Black, 5 = Other/Multi 
    race_label = dplyr::case_when(
      race_ethnicity == 1 ~ "Mexican American",
      race_ethnicity == 2 ~ "Other Hispanic",
      race_ethnicity == 3 ~ "Non-Hispanic White",
      race_ethnicity == 4 ~ "Non-Hispanic Black",
      race_ethnicity == 5 ~ "Other/Multi"))

#sex proportionality summary per dataset 
sex_summary <- demographic_comparison %>% 
  dplyr::count(dataset, 
               sex_label, 
               name = "count") %>% 
  dplyr::mutate(
    proportion = count / sum(count),
    .by = dataset)

#race/ethnicity proportional summary per dataset 
race_summary <- demographic_comparison %>% 
  dplyr::filter(!is.na(race_label)) %>% #need to drop any NA race rows 
  dplyr::count(dataset, 
               race_label, 
               name = "count") %>% 
  dplyr::mutate(
    proportion = count / sum(count),
    .by = dataset,
    #order race categories for consistent plotting 
    race_label = forcats::fct_reorder(race_label, proportion, .desc = TRUE))

```

```{r}
#| label: core-vs-biomarker-sex-plot
#| code-fold: true
#| echo: false 
#| output: false

#plot 1: sex comparison 
sex_comparison_plot <- sex_summary %>% 
  ggplot(aes(
    x = sex_label, 
    y = proportion, 
    fill = dataset)) +
  #dodge lets me avoid stacking - would be more than 100% female w skew in 2.2
  geom_col(position = "dodge") + 
  #geom_col() +
  #use related color theme
  scale_fill_viridis_d(
    option = "magma", #related palette
    begin = 0.2, 
    end = 0.6) + #avoid the blinding white end
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1)) +
    #limits = c(0, 1)) +
  labs(
    x = "Sex",
    y = "Proportion of Dataset",
    title = "Sex Distribution: Core Dataset vs Biomarker Subset",
    fill = "Dataset") +
  theme_minimal() +
  #title/legend position wrangling 
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust =0.5))
#may want to finilzie without gridlines - need to think and come back 
  
#view it 
#sex_comparison_plot

print(sex_comparison_plot)

#need to decide if i want to save this indivifduals - if so get ggsave block

```

```{r}
#| label: core-vs-biomarker-demo-plot
#| code-fold: true
#| echo: false 
#| output: false

#plot 2: race/ethnicity comparison 
race_comparison_plot <- race_summary %>% 
  ggplot(aes(
    x = race_label, 
    y = proportion, 
    fill = dataset)) +
  #same as before to avoid stacking
  geom_col(position = "dodge") +
  coord_flip() +     #flip axes b/c long race/ethnicity labels 
  #use consistent theme and scale 
  scale_fill_viridis_d(
    option = "magma", 
    begin = 0.2, 
    end = 0.6) +
  scale_y_continuous(
    #use percent, not number form 
    labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Race/Ethnicity",
    y = "Proportion of Dataset",
    title = "Race/Ethnicity Distribution: Core Dataset vs Biomarker Subset",
    fill = "Dataset") +
  theme_minimal() +
  #title/legend position wrangling 
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))


#view it 
#race_comparison_plot

print(race_comparison_plot)

#need to decide if i want to save this individuals - if so get ggsave block

```

The figure below extends the model distribution comparison initiated in section 2.2.4.4 to include race/ethnicity demographics alongside sex distribution in a multiplot in order to visualize and characterize the full scope of detected selection bias. 

```{r}
#| label: core-vs-biomarker-distr-comparison-plot
#| code-fold: true
#| output: true

#multiplot - want to keep original plots independent from multi-plot formatting
#so I can use them in powerpoint  
#step 1: 2 panels/plots stacked vertically 
graph_stack <- (
  (sex_comparison_plot +
     #remove double elements (titles, axes titles, etc) 
     theme(
       plot.title = element_blank(),
       axis.title.y = element_blank(),
       axis.title.x = element_blank())) /
    (race_comparison_plot +
       theme(
         plot.title = element_blank(),
         axis.title.y = element_blank()))
)

#step 2: compose final multiplot w/ shared legend and overall title 
demographic_multiplot <- graph_stack +
  plot_layout(
    guides = "collect",
    heights = c(1, 1)) +
  plot_annotation(
    title = "Demographic Composition: \n Core Dataset vs Biomarker Complete-Case Subset",
    subtitle = "Top: Sex   |   Bottom: Race/Ethnicity",
    theme = theme(
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5))) &
  theme(legend.position = "bottom",
        legend.title = element_blank())

#show
demographic_multiplot

```
```{r}
#| label: demo-distr-multiplot-save
#| echo: false
#| output: false

#save plot - ggplot not flextable syntax 
ggsave(
  filename = here::here("plots", "part2", "demo-distr-multiplot.png"),
  plot = demographic_multiplot,
  dpi = 400,
  width = 10,
  height = 7,
  units = "in",
  bg = "white") #was saving grey - this fixed it
```


This figure visually confirms the sex skew identified in section 2.2.4.4. It also indicates that among the populations represented in the initial cohorts, the biomarker subset retains roughly proportional representation for most populations. The only groups that appear to deviate from proper representation are Non-Hispanic Black and White individuals, facing over- and under- representation, respectively, each at an approximately 5%.  

While this figure provided clarification on racial and sex-based distributions, it did not indicate any potential sources of the sex skew. Sex skew was not a direct consequence of the source filing mechanisms unveiled in section 2.3.2.1, but beyond that exclusion, the source of the skew is unknown. Further investigation is encouraged, but could not be applied at this time due to project timeline constraints. Regardless of source, the established sex effect is large enough that the biomarker model's interpretation of any sex-related feature is compromised, including the unexpected negative coefficient for female status that the logistic regression model supplied.

It is important to note that while this figure reflects the accurate preservation of racial/ethnic composition of the initial data, that does not mean that NHANES adequately captured sufficient and representative data of all major ethnic groups, as Asian, Middle Eastern/Arab, Oceanian/Pacific Islander, and Native populations are notably missing. As discussed in section 1.4.3, representation bias does limit the generalizability of the findings, and further investigation should include more representative sampling. 

### <u> 2.3.3: Biomarker Distributions Across HSI Categories </u>

##### <u> 2.3.3.1: Distribution Panel </u>

This section explores biomarker distributions across the three HSI categories to assess individual biomarkers for meaningful distribution separation that aligns with predictive signaling behavior observed in section 2.2. 

```{r}
#| label: biomarker-distribution-panel
#| code-fold: true
#| output: true

# BIOMARKER DISTRIBUTIONS ACROSS HSI CATEGORIES -  -------------
#
#violin + sina plot panel of 8 novel biomarkers faceted across healthy/borderline/nafld_like
#applied to biomarker subset so all rows have complete biomaarker data 

#first need to pivot biomarkers into long format for facetting 
#same reshape pattern as PLSC476 demographic_table 
#
biomarker_long <- biomarker_subset %>% 
  dplyr::filter(!is.na(HSI_category)) %>%    #drop any HSI NA rows 
  dplyr::select(HSI_category, 
                homocysteine, 
                serum_folate, 
                rbc_folate, 
                crp,
                mm_acid, 
                ggt, 
                vit_b12, 
                ferritin) %>% 
  #pivot longer so biomarkers become a variable 
  tidyr::pivot_longer(
    cols = -HSI_category, 
    names_to = "biomarker",
    values_to = "value") %>% 
  dplyr::mutate(
    #rename biomarkers for cleaner display 
    biomarker = dplyr::recode(
      biomarker,
      "homocysteine" = "Homocysteine",
      "serum_folate" = "Serum Folate",
      "rbc_folate" = "RBC Folate",
      "crp" = "CRP",
      "mm_acid" = "MMA",
      "ggt" = "GGT",
      "vit_b12" = "Vitamin B12",
      "ferritin" = "Ferritin"))


#now plot - violin + sina layered, faceted by biomarker 
#same pattern as hw3 PLOT E (violin + sina) 
biomarker_distribution_plot <- 
  ggplot(biomarker_long, aes(x = HSI_category,
                             y = value,
                             color = HSI_category,
                             fill = HSI_category)) +
  geom_violin(alpha = 0) +
  ggforce::geom_sina(alpha = 0.4) +
  #log transform y-axis for skewed biomarker distributions - from Unit 07.A
  scale_y_continuous(trans = "log10") +
  #same color theme as PLSC476 - need both color and fill since mapped in aes
  scale_color_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) +
  scale_fill_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) + #avoid blinding yellow end 
  #facet by biomarker - free_y lets each biomarker have its own scale range 
  #scaling options used in class did not work
  # - had to look this one up but seems to address variation okay 
  facet_wrap(~biomarker, 
             scales = "free_y",
             ncol = 4) +
  labs(
    x = NULL,
    #x = "HSI Category",
    color = "HSI Category",
    fill = "HSI Category", #now show in legend
    y = "Biomarker Value (log10 scale)",
    title = "Novel Biomarker Distributions \n Across HSI Severity Categories") +
  theme_minimal() +
  #title/legend position wrangling
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),     #hide x tick labels - redundant w/ legend
        axis.ticks.x = element_blank(),    #hide x tick marks too 
        plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank())

#view it 
biomarker_distribution_plot
```

```{r}
#| label: hsi-biomarker-multiplot-save
#| echo: false
#| output: false

#save plot - ggplot not flextable syntax 
ggsave(
  filename = here::here("plots", "part2", "biomarker-by-HSI-multiplot.png"),
  plot = biomarker_distribution_plot,
  dpi = 400,
  width = 10,
  height = 7,
  units = "in",
  bg = "white") #was saving grey - this fixed it
```

##### <u> 2.3.3.2: Biomarker Interpretation </u>

This figure shows how each biomarker trends across HSI categories, with each biomarker's y-axis scaled to it's respective value range and log transformed for consistency. Multiple features, including homocysteine, GGT, folate metrics, CRP, and B12 display possible outliers that may impact distribution trends. However, outlier identification and filtering could not be applied at this time due to project timeline constraints. Further investigation into outliers is heavily encouraged in follow up studies. 

- need to ask jeff before finalizing - wish i had more time for filtering outliers and re-running but i do not 

- ask to confirm trends: 

I see: 

1) CRP - strong directional shift (+ association w/ nafld)
  
   - shape of distribution changes notably across bins - only feature that shows this (bottom to top heavy density movement)
    - theres one line at the bottom for healthy, and borderline, but theres a second floor above it too? and a 3rd? around 0.2 and 0.5 respectively what are those?
   - matches expectations established in section 2.2 - means CRP is part of the evidence that the model's coefficients are correct where they're expected and incorrect (likely due to multicollinearity) where they contradict expected trends (aka females)
   
2) Ferritin -  moderate directional shift (+ association w/ nafld)

- as proximity to NAFLD-like states increases, so does the variability of ferrin values - distribution range keeps expanding across bins? what would cause that?

 - possible explanation - ferritin is used as an iron storage marker and an acute inflammation marker - elevated iron isnt a uniform signal of FLD so maybe this shows the variation (which also points to the fact that NAFLD is an aggregation of symptoms, not a uniform diagnosis based on a singular shared mechanism)
  
- model gave ~0 coefficient - wasn't indicated as a meaningful predictor but visual contradicts that somewhat - tracks with severity but weaker signal - could be caused by mechanistic distance, shared signal with other features or something else 

3) GGT - moderate-small directional trend (+ association w/ nafld)

- borderline and healthy classes look more similar in terms of mode value
- showing similar distribution creep/range expansion with increasing severity as ferritin 
  - potential source? same as ferritin - will probably discuss together 
- most significant outliers look to be healthy? 

4) Hcy - moderate directional trend (+ association w/ nafld)

   - shift mostly seen through density/mode value 
   - contradicts model coefficient findings but aligns with mechanistic expectations - contradiction supports the multicollinearity/suppression hypothesis from 2.2.4.4 — biomarker has correct directional signal at distribution level, but model coefficient is distorted by shared variance with other features

5) MMA - unclear/no trend 

  - no density, range, or directional trend readily apparent
  - contradicts negative ML findings 
  - interpretation - unclear-  could be artifact/reflect model conditioning

6) RBC - small/no trend

  - ML reported negative relationship (biologically expected)
  - no clear visual trend - discuss with MMA 

7) serum - small trend - negative

  - negative aligns with biological mechanism + ML portion 
  - distribution shape consistent, but negative trend apparent in decreasing floor value per cohort aligns with expectation - though without outlier evaluation, may not be assertable 
  - trend small but more visually apparent than RBC - weird bc RBC is more stable/long term, so wed expect to see any impact present there 
  - possible attributions: noise from data, or biologically significant (AKA NALFD acute folate shows more variety - aligns with increased range present in other, stronger deterministic factors/biomarkers)

8) B12 - small trend - negative

  - visually stronger than ML portion indicated (like ferritin)
  - negative trend aligns with biological expectation (B12 deficiency impairs methylation/is associated with metabolic dysfunction)

### <u> 2.3.4: Sex-Stratified View of Biomarker Subset </u>

##### <u> 2.3.4.1: NAFLD Class Counts by Sex </u>

Section 2.3.2.2 confirmed a sex skew in the biomarker subset and noted that any sex-related feature conclusions are likely compromised. This section addresses the downstream analytical impact of that skew by stratifying NAFLD class distributions and selected biomarker patterns by sex. Doing this isolates biomarker behavior within sex-specific samples, so that sex no longer functions as a confounding variable in the biomarker subset. 

```{r}
#| label: nafld-by-sex-biomarker-subset-prep
#| code-fold: true
#| output: true

# NAFLD CLASS COUNTS BY SEX IN BIOMARKER SUBSET -  -------------
#
#grouped bar chart showing NAFLD +/- counts broken down by sex 
#within the biomarker subset to visualize the inverted prevalence flagged in 2.2.4.4 

#summary of NAFLD counts by sex 
nafld_sex_summary <- biomarker_subset %>% 
  dplyr::filter(!is.na(NAFLD),
                !is.na(sex)) %>% 
  dplyr::mutate(
    #recode sex for readability 
    sex_label = dplyr::case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female"),
    #recode NAFLD for readability 
    nafld_label = dplyr::case_when(
      NAFLD == 1 ~ "NAFLD-positive",
      NAFLD == 0 ~ "Healthy")) %>% 
  dplyr::count(sex_label, 
               nafld_label, 
               name = "count") %>% 
  dplyr::mutate(
    #proportion within sex so bars are comparable across sex groups 
    proportion = count / sum(count),
    .by = sex_label)

```


```{r}
#| label: nafld-by-sex-biomarker-subset-plot
#| code-fold: true
#| output: true

#okay now I can plot 
nafld_sex_plot <- 
  ggplot(nafld_sex_summary, aes(x = sex_label,
                                y = proportion,
                                fill = nafld_label)) +
  geom_col(position = "dodge") + 
  scale_fill_viridis_d(
    option = "magma",
    begin = 0.2,
    end = 0.6) + #avoid blinding white end 
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Sex",
    y = "Proportion within Sex",
    title = "NAFLD Class Distribution by Sex \n in Biomarker Subset",
    fill = "NAFLD Status") +
  theme_minimal() +
  #title/legend position wrangling 
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank())

#view it 
nafld_sex_plot

```
```{r}
#| label: nafld-sex-plot-save
#| echo: false
#| output: false

#save plot 
ggsave(
  filename = here::here("plots", "part2", "nafld_by_sex_biomarker_subset.png"),
  plot = nafld_sex_plot,
  dpi = 400,
  width = 12,
  height = 7,
  units = "in",
  bg = "white")
```

This figure visually confirms the inverted NAFLD prevalence noted in section 2.2.4.4, in which biomarker subset males show roughly equal class balance while females skew towards healthy classification despite making up 89% of the subset. This now confirmed pattern is what compounded interpretation difficulties in the biomarker model, since not only was the model trained on heavily female skewed-data, but those females were disproportionately healthy. Stratification removes the confounding sex aspect entirely. While the female sample is still skewed towards healthy classification, and therefor limited in NAFLD-positive analysis in females specifically, female distributions can still be used for baseline sex-difference observations and as a consistency check against male-specific findings. 

##### <u> 2.3.4.2: Homocysteine by Sex & HSI Category </u>

```{r}
#| label: homocysteine-by-sex-hsi
#| code-fold: true
#| output: true

# HOMOCYSTEINE DISTRIBUTION BY SEX AND HSI CATEGORY -  -------------
#
#violin + sina of homocysteine across HSI categories, faceted by sex 
#tests whether Hcy pattern differs between sexes - ties to MTHFR pathway narrative 

#homocysteine long format w/ sex labels - prep but small so same chunk 
hcy_sex_data <- biomarker_subset %>% 
  #get rid of any na skew 
  dplyr::filter(!is.na(HSI_category),
                !is.na(sex),
                !is.na(homocysteine)) %>% 
  dplyr::mutate(
    sex_label = dplyr::case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female"))


#plot - violin + sina pattern same as 2.3.3, but this time faceted by sex - 
#need to evestigate results of previous chunk further 
hcy_sex_plot <- 
  ggplot(hcy_sex_data, aes(x = HSI_category,
                           y = homocysteine,
                           color = HSI_category,
                           fill = HSI_category)) +
  geom_violin(alpha = 0) +
  ggforce::geom_sina(alpha = 0.5) +
  #log transform since homocysteine is right-skewed 
  scale_y_continuous(trans = "log10") +
  scale_color_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) +
  scale_fill_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) + 
  #facet by sex 
  facet_wrap(~sex_label, ncol = 2) + 
  labs(
    x = NULL, 
    y = "Homocysteine (log10 scale)",
    title = "Homocysteine Distribution by Sex \n Across HSI Severity Categories",
    fill = "HSI Category",
    color = "HSI Category") +
  theme_minimal() +
  #title/legend position wrangling 
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank())

#view it 
hcy_sex_plot
```
```{r}
#| label: HSI-sex-homocysteine-plot-save
#| echo: false
#| output: false

#save plot 
ggsave(
  filename = here::here("plots", "part2", "homocysteine_by_sex_hsi.png"),
  plot = hcy_sex_plot,
  dpi = 400,
  width = 12,
  height = 7,
  units = "in",
  bg = "white")
```




notes: - immediately can see sex based population skew - males seem to exhibit higher homocysteine - not explicitely stated this project, bc uses mouse model but ANSC project paper notes that men seen to use BHMT pathway to compensate - so why increased hcy? bc women are skewed healthier? - wish i had MTHFR data - also makes me question applicability of stats to gen pop, esp with women seemingly more healthy - doesnt align with known data, need to verify. -worthy of discussion - meaningful differences - should proceed with secondary analysis of most variable features (ggt, crp)

##### <u> 2.3.4.3: CRP & GGT by Sex & HSI Category </u>

```{r}
#| label: crp-ggt-by-sex-hsi
#| code-fold: true
#| output: true

# CRP & GGT DISTRIBUTIONS BY SEX AND HSI CATEGORY -  -------------

#pivot CRP & GGT into long format w/ sex labels  
crp_ggt_sex_data <- biomarker_subset %>% 
  dplyr::filter(!is.na(HSI_category),
                !is.na(sex),
                !is.na(crp),
                !is.na(ggt)) %>% 
  dplyr::mutate(
    sex_label = dplyr::case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female")) %>% 
  dplyr::select(HSI_category, 
                sex_label, 
                crp, 
                ggt) %>% 
  #pivot longer so biomarkers become a variable 
  tidyr::pivot_longer(
    cols = c(crp, ggt), 
    names_to = "biomarker",
    values_to = "value") %>% 
  dplyr::mutate(
    #rename for cleaner display 
    biomarker = dplyr::recode(
      biomarker,
      "crp" = "CRP",
      "ggt" = "GGT"))


#plot - violin + sina, faceted by biomarker (rows) x sex (cols) 
crp_ggt_sex_plot <- 
  ggplot(crp_ggt_sex_data, aes(x = HSI_category,
                               y = value,
                               color = HSI_category,
                               fill = HSI_category)) +
  geom_violin(alpha = 0) +
  ggforce::geom_sina(alpha = 0.5) +
  #log transform y-axis for skewed biomarker distributions
  scale_y_continuous(trans = "log10") +
  scale_color_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) +
  scale_fill_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) + 
  #facet grid - biomarker rows x sex columns; free_y lets each biomarker have its own scale 
  facet_grid(biomarker ~ sex_label, 
             scales = "free_y") +
  labs(
    x = NULL,
    y = "Biomarker Value (log10 scale)",
    title = "CRP & GGT Distributions by Sex \n Across HSI Severity Categories",
    fill = "HSI Category",
    color = "HSI Category") +
  theme_minimal() +
  #title/legend position wrangling 
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))


#save plot 
#ggsave(
  #filename = here::here("plots", "part2", "crp_ggt_by_sex_hsi.png"),
  #plot = crp_ggt_sex_plot,
  #dpi = 400,
  #width = 12,
  #height = 9,
  #units = "in",
  #bg = "white")

#view it 
crp_ggt_sex_plot

#ask jeff opinion on flipping order axes vs label for y 
#- i know its not conventions but i felt like i naturally looked for the labels first, then the measurements 

```

```{r}
#| label: crp-ggt-by-sex-hsi-drafting
#| code-fold: true
#| output: true

# CRP & GGT DISTRIBUTIONS BY SEX AND HSI CATEGORY -  -------------

#pivot CRP & GGT into long format w/ sex labels  
crp_ggt_sex_data <- biomarker_subset %>% 
  dplyr::filter(!is.na(HSI_category),
                !is.na(sex),
                !is.na(crp),
                !is.na(ggt)) %>% 
  dplyr::mutate(
    sex_label = dplyr::case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female")) %>% 
  dplyr::select(HSI_category, 
                sex_label, 
                crp, 
                ggt) %>% 
  #pivot longer so biomarkers become a variable 
  tidyr::pivot_longer(
    cols = c(crp, ggt), 
    names_to = "biomarker",
    values_to = "value") %>% 
  dplyr::mutate(
    #rename for cleaner display 
    biomarker = dplyr::recode(
      biomarker,
      "crp" = "CRP",
      "ggt" = "GGT"))


#plot - violin + sina, faceted by biomarker (rows) x sex (cols) 
crp_ggt_sex_plot <- 
  ggplot(crp_ggt_sex_data, aes(x = HSI_category,
                               y = value,
                               color = HSI_category,
                               fill = HSI_category)) +
  geom_violin(alpha = 0) +
  ggforce::geom_sina(alpha = 0.5) +
  #log transform y-axis for skewed biomarker distributions
  scale_y_continuous(trans = "log10") +
  scale_color_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) +
  scale_fill_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) + 
  #facet grid - biomarker rows x sex columns; free_y lets each biomarker have its own scale 
  #ask jeff opinion on flipping order axes vs label for y 
#- i know its not conventions but i felt like i naturally looked for the labels first, then the measurements 
#he agreed that we could flip sex and biomarker to improve visual intiutiveness - coord_flip? - nope - just need to flip orientation of ~ line 
  facet_grid(sex_label ~ biomarker, 
             scales = "free_y") +
  labs(
    x = NULL,
    y = "Biomarker Value (log10 scale)",
    title = "CRP & GGT Distributions by Sex \n Across HSI Severity Categories",
    fill = "HSI Category",
    color = "HSI Category") +
  theme_minimal() +
  #title/legend position wrangling 
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))


#save plot 
#ggsave(
  #filename = here::here("plots", "part2", "crp_ggt_by_sex_hsi.png"),
  #plot = crp_ggt_sex_plot,
  #dpi = 400,
  #width = 12,
  #height = 9,
  #units = "in",
  #bg = "white")

#view it 
crp_ggt_sex_plot

```

```{r}
#| label: crp-ggt-by-sex-hsi-drafting-2
#| code-fold: true
#| output: true

# CRP & GGT DISTRIBUTIONS BY SEX AND HSI CATEGORY -  -------------
##idea for label/readability issue - can't risk compressed data but can switch category legend and sex label positions - more inteutive and better contexual info for valuable real estate under caption - benefit of reducing visual importance of sex label and balancing labels a little better than og

#pivot CRP & GGT into long format w/ sex labels  
crp_ggt_sex_data <- biomarker_subset %>% 
  dplyr::filter(!is.na(HSI_category),
                !is.na(sex),
                !is.na(crp),
                !is.na(ggt)) %>% 
  dplyr::mutate(
    sex_label = dplyr::case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female")) %>% 
  dplyr::select(HSI_category, 
                sex_label, 
                crp, 
                ggt) %>% 
  #pivot longer so biomarkers become a variable 
  tidyr::pivot_longer(
    cols = c(crp, ggt), 
    names_to = "biomarker",
    values_to = "value") %>% 
  dplyr::mutate(
    #rename for cleaner display 
    biomarker = dplyr::recode(
      biomarker,
      "crp" = "CRP",
      "ggt" = "GGT"))


#plot - violin + sina, faceted by biomarker (rows) x sex (cols) 
crp_ggt_sex_plot <- 
  ggplot(crp_ggt_sex_data, aes(x = HSI_category,
                               y = value,
                               color = HSI_category,
                               fill = HSI_category)) +
  geom_violin(alpha = 0) +
  ggforce::geom_sina(alpha = 0.5) +
  #log transform y-axis for skewed biomarker distributions
  scale_y_continuous(trans = "log10") +
  scale_color_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) +
  scale_fill_viridis_d(
    option = "plasma", 
    begin = 0, 
    end = 0.7) + 
  #facet grid - biomarker rows x sex columns; free_y lets each biomarker have its own scale 
  #ask jeff opinion on flipping order axes vs label for y 
#- i know its not conventions but i felt like i naturally looked for the labels first, then the measurements 
#he agreed that we could flip sex and biomarker to improve visual intiutiveness - coord_flip? - nope - just need to flip orientation of ~ line 
       
  facet_grid(biomarker ~ sex_label,
             scales = "free_y",
             #switching sex label position here 
             switch = "x") +
  labs(
    x = NULL,
    y = "Biomarker Value (log10 scale)",
    #needed to move split but better now 
    title = "CRP & GGT Distributions by \n Sex Across HSI Severity Categories",
    fill = "HSI Category",
    color = "HSI Category") +
  theme_minimal() +
  #title/legend position wrangling - switching legend position here 
  theme(legend.position = "top",
        #legend.direction("horizontal,")
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))


#save plot 
#ggsave(
  #filename = here::here("plots", "part2", "crp_ggt_by_sex_hsi.png"),
  #plot = crp_ggt_sex_plot,
  #dpi = 400,
  #width = 12,
  #height = 9,
  #units = "in",
  #bg = "white")

#view it 
crp_ggt_sex_plot

```
Future analysis should address these potential outliers, but due to project timeline constraints, this could not be further explored at this time.

notes:

-   also displays sex difference, especially the healthy class
-   need to look for sex changes
-   whats going on witht the linear trend for CRP females low end
-   dont forget outlier removal on next pass - maybe will do upstream all at once since seems to be an issue repeatedly

### <u> 2.3.5: Summary & Integration with Part 2.2 Findings </u>

These diagnostic observations refine, but do not replace, the interpretive framing carried forward from Part 1. The extent to which MTHFR-pathway biomarker patterns observed in Part 1 map onto population-level FLD signal is taken up in the integration discussion that follows.

- blurb to remember: "The cross-language reproducibility approach (re-deriving HSI labels in R from raw NHANES variables rather than importing the Python output) initially surfaced a NaN-handling discrepancy in the Python label derivation that affected reported category counts but not model training results. The discrepancy was corrected before downstream analysis. Independent re-derivation in a second language is not a standard step in undergraduate capstone work, but in this project it provided meaningful verification that label-construction logic matched across implementations."
